A global sensitivity analysis of model uncertainty in aeroelastic wind turbine models

### 2020-02-20 Papers/Torque_Paper/content.tex
minor updates
shortened description of PCE, several cleanups

The goal of this work is to perform a systematic assessment of the uncertainty in model parameters in BEM models. We approach this by performing a global sensitivity analysis based on the Sobol expansion approach, which decomposes the total variance of the quantity of interest (model output) into contributions from individual parameters and their combinations, similar to \cite{Echeverria2017,Murcia2018, Rinker2016a}. We will employ the uncertainty quantification toolbox \texttt{UQLab} \cite{uqlab}, which computes the Sobol indices based on a sparse polynomial chaos expansion.

The main novelty of the work lies in the study of the effect of model uncertainties rather than external conditions or geometric uncertainties. As such, the current sensitivity analysis is part of the so-called WindTrue project, in which the long-term goal is the development of calibrated BEM models, that possess a quantified level of uncertainty. Anticipating on this goal, we are using the NM80 wind turbine model from the Tjaereborg wind farm, for which extensive measurement data is available through the DanAero project \cite{Troldborg2013} and IEA Task 29. The measurement data will enable the calibration step, and our sensitivity analysis will make clear which of the model parameters should be used in this calibration step.

@@

\section{Aeroelastic model uncertainties}\label{sec:model_description}
%\subsection{BEM model uncertainties}
State-of-the-art wind-turbine models simulate the aeroelastic behavior of wind turbines by combining the concept of momentum conservation of the flow (BEM theory) with the equations of motion for the structure, possibly extended with the hydrodynamics of the sea and control algorithms \cite{Vorpahl2013}. In this work, we only concentrate on the first aspect, namely the prediction of flow and blade forces as given by the BEM method. The particular BEM code that we use is ECN's \texttt{Aero-Module} \cite{Boorsma2012}.

For the case of a rigid turbine, with a uniform inflow field, some of the main uncertainties that arise in the BEM formulation are \cite{Hansen1993}:
\begin{itemize}
\item The use of (2D) \textit{airfoil polars}; uncertainty arises not only because the actual flow along the blade is 3D, but also because the 2D polars itself can be inaccurate, either when obtained from measurements or 2D codes like \texttt{XFOIL};
%\item Sheared inflow
\item The assumption of locally 2D flow along the blade is also invalid at the tip; this is typically corrected via \textit{a tip loss model};
\item Three-dimensional flow effects also cause stall delay at the root; this is typically modeled via \textit{a dynamic stall model} and associated time constant;
\item Yawed inflow is not naturally included and typically incorporated via \textit{a wake correction}.
%\item Dynamic wake - Oye time lag constant
%\item Tower shadow constants " Potential flow / Downwind Empirical model
\end{itemize}
Next to these model uncertainties we will consider geometric uncertainties: the chord and twist distribution along the blade.

@@

%\subsection{Geometric uncertainties with NURBS curves}
The chord, twist, lift coefficient, and drag coefficient are functions of the radial distance along the blade, leading to a very high-dimensional number of uncertain parameters. In order to reduce this number, we parametrize the variation along the blade by using Non-Uniform Rational Basis Splines (NURBS) \cite{rogers2000}. In this way only a limited number of control points is needed to approximate a large variety of curves, so that the resulting number of uncertain parameters is relatively small. 
%To perturb the geometrical parameters of the turbine blade, such as the reference chord or twist. This parameterization will also be used to perturb lift-and drag polars. %Further, the set of control points and knots can be directly manipulated to control the smoothness and curvature.
NURBS have already been exploited in several aerodynamics applications, for instance, design optimization of wind-turbine blades, e.g.~\cite{Bottasso2014,Ribeiro2012}, and to parametrize chord and twist curves \cite{Echeverria2017}. In the following, we briefly outline the procedure used to parametrize a curve using NURBS.

@@

In order to parameterize a given curve, e.g.\ the blade reference chord, we choose a degree $p$ and a number of knots $X =\{r_j\}_{j=1}^n$. The knots can be uniformly spaced or chosen heuristically (for advanced knot selection approaches for knot selection see \cite{Li2005}). Next we sample $S_{ref}(r)$ at $X$ and also compute $B_{i,p}(r_j), \text{ for }i, j = 1,2, ..., n$. We then formulate a linear system and solve for the set of control points $\mathbf{c}=\{c_i\}_{i=1}^n$ as:
\begin{equation}\label{nurbs_inversion}
\mathbf{B}\mathbf{c} = \mathbf{S}_{ref},
\end{equation}
where $\mathbf{S}_{ref}\in\mathbb{R}^{n}$ is a vector containing sampled values of $S_{ref}(r)$ and $\mathbf{B}\in \mathbb{R}^{n\times n}$ is a matrix with the $j$-th row consisting of values of $n$ B-splines sampled at location $r_j$. Once the control points are obtained, the approximate reference curve $S_n(r)$ follows from \eqref{NURB_curve}. Fig.~\ref{nurbs_approx} shows a parameterization of the chord curve using $n=9$ control points and second-order ($p=2$) B-splines.
\begin{figure}[h!]
\centering
\includegraphics[trim={0cm 0cm 0 0cm},clip, scale=1.3]{nurbs_approx.eps}
\caption{Parametrization of chord using 9 control points and second-order B-splines.}
\label{nurbs_approx}
\end{figure}

@@

\subsection{Sobol analysis}
We employ the variance-based Sobol decomposition to perform global sensitivity analysis. This approach allows for quantification of the relative importance of the input parameters on a scale of $[0,1]$ known as \emph{Sobol indices}. In a Sobol analysis, we express the total variance of the output in term of contributions from individual parameters and their combinations. Consider a generic BEM model $f(\mathbf{z})$, depending on uncertain parameters $\mathbf{z}$ and returning a quantity of interest $Y$ (e.g.\ power, loads, moments):
\begin{equation}\label{nonlinear_model}
Y = f({\mathbf{z}}),
\end{equation}
where ${\mathbf{z}} = [z_1, z_2, ..., z_M]\in \mathcal{D}_{\mathbf{z}}\in \mathbb{R}^M$ is a vector of random variables (parameters). For ease of notation, we assume these input parameters to be uniformly distributed i.e. $z_i \sim \mathcal{U}(0,1)$ and the support of the input set is $\mathcal{D}_{\mathbf{z}} = [0,1]^M$ where $M$ is the total number of input parameters. The Sobol decomposition is based on the following hierarchical representation \cite{RSmith}:
\begin{equation}\label{sobol_decomp}
f(z_1, z_2, ..., z_M) = f_0+\sum_{i=1}^M f_i(z_i) + \sum_{1\leq i<j\leq M} f_{ij}(z_i,z_j) + ... + f_{1, 2, ..., M}(z_1, z_2, ..., z_M).
\end{equation}
Here the zeroth-order function $f_0$ is the mean response of $f$, the first-order univariate functions $f_i(z_i)$ quantify independent contributions due to each random input, and second-order functions $f_{i,j}(z_i,z_j)$ represent the effect of interaction between $z_i$ and $z_j$ on the response $f$. Formally, the zeroth-, first- and second-order terms are defined as:
\begin{align}\label{sobol_terms}
f_0 &= \int_{\mathcal{D}_{\mathbf{z}}}f(\mathbf{z})d\mathbf{z},\\
f_i(z_i) &= \int_{{\mathcal{D}_{\mathbf{z}}}^{M-1}}f(\mathbf{z})d\mathbf{z}_{\sim \{i\}} - f_0,\\
f_{ij}(z_i,z_j) &= \int_{{\mathcal{D}_{\mathbf{z}}}^{M-2}}f(\mathbf{z})d\mathbf{z}_{\sim \{i,j\}} - f_i(z_i) - f_j(z_j) - f_0,
\end{align}
where ${\mathcal{D}_{\mathbf{z}}}^{M-1} = [0,1]^{M-1}$ and the notation $\mathbf{z}_{\sim \{i,j\}}$ denotes the vector having all the components of $\mathbf{z}$ except in the set $\{i,j\}$. Higher-order terms are interpreted in a similar manner. Note that \eqref{sobol_decomp} is only valid for independent input parameters. We only consider a second-order expansion of \eqref{sobol_decomp}, as for many applications this is sufficient because higher-order terms have negligible effect on the output response.

As an example, $Y$ can represent the power output of a wind turbine, with random inputs windspeed $(z_1)$, wind standard deviation $(z_2)$ and rotational speed $(z_3)$. $f_0$ then represents the mean power output considering all the three random inputs, $f_1(z_1)$ represent the independent contribution of windspeed on the power and $f_{1,2}(z_1,z_2)$ quantifies the interactions of wind speed and wind standard deviation on power.

The definition of the variance is:
\begin{equation}\label{tot_var}
D = \text{Var}[f(\mathbf{z})] = \int_{\mathcal{D}_{\mathbf{z}}} f^2(\mathbf{z})d\mathbf{z} - f_0^2.
\end{equation}
As the expansion in \eqref{sobol_decomp} is not unique, some orthogonality conditions must be imposed  \cite{Rabitz1999,SOBOL2001271} which allows to express the total variance as
\begin{equation}
D = \sum_{i=1}^M D_i + \sum_{1\leq i<j\leq M} D_{ij},
\end{equation}
where the first- and second-order partial variances are defined via the Sobol decomposition as
\begin{align}
D_i &= \int_{0}^{1} f^2_i(z_i)dz_i, & D_{ij} &= \int_{0}^{1} \int_{0}^{1} f^2_{ij}(z_i,z_j)dz_idz_j \label{partial_variance2}. 
\end{align}
The first- and second-order Sobol indices are then computed as
\begin{equation}\label{sobol_ind}
S_i = \frac{D_i}{D}, \quad S_{ij} = \frac{D_{ij}}{D}, \quad i,j=1,2, ..., M.
\end{equation}
%and the sum of all indices satisfy
%\begin{equation}
%\sum_{i=1}^M S_i + \sum_{1\leq i<j\leq M} S_{ij} = 1.
%\end{equation}
The total effect of $z_i$ on the output $Y$ can be quantified using the total sensitivity index
\begin{equation}
S_{T_i} = S_i + \sum_{j=1}^M S_{ij}.
\end{equation}
The total sensitivity indices can be interpreted as an importance measure for the parameter $z_i$, therefore a large $S_{T_i}$ implies that $z_i$ has a strong influence on $Y$. The computation of $S_{T_i}$ requires the approximation of partial variances defined in \eqref{partial_variance2}. Using the Monte Carlo method for computing these variances can be prohibitive for computationally expensive models $Y$. On the other hand, Polynomial Chaos Expansions (PCE) based approaches provide a more efficient alternative for sensitivity analysis, and will be used here. As shown in \cite{SUDRET2008964}, one can compute the Sobol indices analytically by post-processing the PCE coefficients.

\subsection{PCE-based Sobol indices computation}
The PCE approximation $f^{K}(\mathbf{z})$ of the computational model $Y$ can be defined as a weighted sum of multivariate polynomials in $\mathbf{z}$ \cite{RSmith}
\begin{equation}\label{eq:PCE}
Y= f(\mathbf{z}) \approx f^{K}(\mathbf{z}) = \sum_{|\mathbf{k}| = 0}^K w_{\mathbf{k}}\Psi_{\mathbf{k}}(\mathbf{z}),
\end{equation}
where $\mathbf{k}\in \mathbb{N}_0^M$ is a $M$-dimensional multi-index with magnitude $|\mathbf{k}| = k_1+k_2 + ... + k_M$ and $\Psi_{\mathbf{k}}(\mathbf{z})$ a multivariate polynomial basis, which is computed using a tensor product of univariate polynomials $\psi_{k_i}^{(i)}(z_i)$
\begin{equation}
\Psi_{\mathbf{k}}(\mathbf{z}) := \prod_{i=1}^M\psi_{k_i}^{(i)}(z_i),
\end{equation}
where $k_i$ is the order of the univariate polynomial. The choice of the univariate orthogonal polynomial $\psi_{k_i}$ depends on the type of random variable, $z_i$. For example, for uniform random variables the Legendre family of polynomials is used \cite{Xiu2002}. 
%The multivariate polynomials $\Psi_{\mathbf{k}}(\mathbf{z})$ is orthogonal with respect to the joint probability distribution of random vector $\mathbf{z}$. 
The total number of terms $K$ in \eqref{eq:PCE} can be based for example on a standard truncation scheme in which all polynomials in $M$ random variable with order less than or equal to $p$ are used. In this work, a more advanced truncation schemes known as the hyperbolic truncation scheme is used \cite{BlatmanThesis}. 
%\textcolor{red}{In this work, we use the hyperbolic truncation scheme with a $qnorm = 0.75$.}

The PCE coefficients $w_{\mathbf{k}}$ corresponding to $\Psi_{\mathbf{k}}(\mathbf{z})$ can be computed using a pseudo-spectral approach by using quadrature methods. For a small number of dimensions this is an efficient approach. However, in the current work we will encounter more than 10 dimensions and quadrature methods are typically too expensive as they suffer strongly from the curse of dimensionality. We therefore focus on methods in which the sampling points can be chosen more freely: Ordinary Least Squares (OLS) and Least-Angle Regression (LARS).
%In the current work where we have a limited computational budget, methods that minimizes the approximation error for a given number of samples are more practical. Next, we briefly describe two such methods to compute PCE coefficients based on a fixed number of model evaluations.

@@

\subsubsection{Ordinary Least Squares}
Instead of using a projection-based method one can also use least squares based approaches to compute coefficients \cite{Berveiller2006}. The PCE expansion \eqref{eq:PCE} can be expressed as a vector-matrix product
\begin{equation}\label{eq:LSM}
Y = f^{K}(\mathbf{z}) +\varepsilon_K = \sum_{|\mathbf{k}| = 0}^K w_{\mathbf{k}}\Psi_{\mathbf{k}}(\mathbf{z}) + \varepsilon_K = \mathbf{w}^T\boldsymbol{\Psi}(\mathbf{z}) +\varepsilon_K,
\end{equation} 
where $\mathbf{w} = \{w_0,w_1,...,w_K\}^T$ is the coefficient vector, $\boldsymbol{\Psi}(\mathbf{z}) = \{\Psi_0(\mathbf{z}), \Psi_1(\mathbf{z}), ..., \Psi_K(\mathbf{z})\}^T$ is the matrix containing multivariate polynomials in $\mathbf{z}$, and $\varepsilon_K$ is the approximation error. The least-squares minimization problem can then be formulated as:
\begin{equation}\label{eq:PCE_OLS}
\mathbf{\overline{w}} = \argmin_{\mathbf{w}} \mathbb{E}\left[\left(\mathbf{w}^T \boldsymbol{\Psi}(\mathbf{z}) - f(\mathbf{z})\right)^2\right]
 \end{equation}
The coefficient vector $\mathbf{\overline{w}}$ can be computed using the Ordinary Least Squares (OLS) method. Given $N$ model evaluations $\mathbf{Y} = \{f(\mathbf{z}_1), f(\mathbf{z}_2), ..., f(\mathbf{z}_N)\}$, we can obtain the coefficients as:
\begin{equation}\label{eq:OLS}
\overline{\mathbf{w}} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{Y},
\end{equation}
where $\mathbf{X}\in \mathbb{R}^{N\times K}$ is the design matrix where $\mathbf{X}_{i,j} = \Psi_{j}(\mathbf{z}_i)$, $i = 1,2, ..., N$ and $j = 0,1,2,..., K$. Although the OLS method allows to compute the coefficients using a limited number of model evaluations, one of the main drawbacks of this approach is that it may result in a PCE model that consists of high-order interaction terms that may lead to overfitting. For many engineering problems, low-order interactions between the input variables are most important. Therefore, the least squares form in \eqref{eq:PCE_OLS} can be regularized such that the minimization results is a low-order sparse PCE model, as discussed next.

\subsubsection{Least Angle Regression}
The LARS algorithm \cite{efron2004} is one of the methods that can be applied to compute sparse PCE models \cite{BlatmanThesis, BLATMAN2011}. The key idea is to find the best set of polynomials from a given candidate set that minimizes the regularized form of \eqref{eq:PCE_OLS}:
\begin{equation}\label{regularized_LS}
\mathbf{\overline{w}} = \argmin_{\mathbf{w}} \mathbb{E}\left[\left(\mathbf{w}^T \boldsymbol{\Psi}(\mathbf{z}) - f(\mathbf{z})\right)^2\right] + \lambda ||\mathbf{w}||_1.
\end{equation}
The LARS algorithm in the context of PCE can be summarized as follows. We start with all PCE coefficients for the candidate polynomials set to zero, and find the polynomial, say $\Psi_1(\mathbf{w})$ most correlated with the model evaluations. This polynomial enters the active set. Next, we take the largest step in the direction of $\Psi_1(\mathbf{w})$ until some other polynomial $\Psi_2(\mathbf{w})$ from the candidate set, has as much correlation with the current residual. The second polynomial is now also added to the active set. In the next step, we move both the coefficients in the active set towards their least-square values until a third polynomial from the candidate set shows as much correlation as the current PCE model based on two polynomials shows with the model evaluations. The number of iterations is given by the minimum of $\{K,N\}$. After every iteration a posteriori error, for instance the Leave-One-Out (LOO) cross-validation error, is computed. The active set with the smallest LOO is then chosen as the best sparse PCE model.

@@

\subsection{Sobol index computation}

@@

\subsection{NM80}
\subsubsection{Operating conditions}
As a reference wind turbine, we consider the 2MW NM80 onshore turbine from the DANAERO project \cite{Troldborg2013} with a blade radius of 38.8 m and a hub height of 57.19 m with the mean wind speed at hub height set as 6.1m/s. Furthermore, the RPM is set at 12.3$/\text{min.}$, the blade pitch angle at 0.15 $\deg$ and the turbine yaw angle is zero.

\subsubsection{Choice of uncertain inputs}
The data for lift ($C_l$) and drag ($C_d$) polars are available at four locations along the blade radius: at 11.87 m, 17.82 m, 28.97 m, and 35.53 m. The lift and drag variables at these sections are numbered Cl1 - Cl4 and Cd1 - Cd4 respectively. The reference value of these polars is obtained from the wind-tunnel experiment with 3D corrections. Random samples of chord, twist, lift- and drag-polars are obtained by perturbing the control points with a uniformly distributed random variable.

>>>>

The goal of this work is to perform a systematic assessment of the uncertainty in model parameters in BEM models. We approach this by performing a global sensitivity analysis based on the Sobol expansion approach, which decomposes the total variance of the quantity of interest (model output) into contributions from individual parameters and their combinations, similar to \cite{Echeverria2017,Murcia2018, Rinker2016a}. We will employ the uncertainty quantification toolbox \texttt{UQLab} \cite{uqlab}, which computes the Sobol indices based on a sparse polynomial chaos expansion. The main novelty of the work lies in creating a generic framework for studying parametric uncertainteis, including both model uncertainties and geometric uncertainties. As such, the current sensitivity analysis is part of the so-called WindTrue project, in which the long-term goal is the development of calibrated BEM models. Anticipating on this goal, we are using the NM80 wind turbine model from the Tjaereborg wind farm, for which extensive measurement data is available through the DanAero project \cite{Troldborg2013} and IEA Task 29. The current sensitivity analysis will make clear which of the model parameters are to be used in the calibration phase.

@@

\section{Aeroelastic model uncertainties}\label{sec:model_description}
%\subsection{BEM model uncertainties}
State-of-the-art wind-turbine models simulate the aeroelastic behavior of wind turbines by combining the concept of momentum conservation of the flow (BEM theory) with the equations of motion for the structure, possibly extended with the hydrodynamics of the sea and control algorithms \cite{Vorpahl2013}. In this work, we concentrate on the first aspect, namely the prediction of flow and blade forces as given by the BEM method. The particular BEM code that we use is ECN's \texttt{Aero-Module} \cite{Boorsma2012}. For the case of a rigid turbine, with a uniform inflow field, the main uncertainties that arise in the BEM formulation are \cite{Hansen1993}:
\begin{itemize}
\item The use of (2D) \textit{airfoil polars}; uncertainty arises not only because the actual flow along the blade is 3D, but also because the 2D polars itself can be inaccurate, either when obtained from measurements or 2D codes like \texttt{XFOIL};
%\item Sheared inflow
\item The assumption of locally 2D flow along the blade is also invalid at the tip; this is typically corrected via \textit{a tip loss model};
\item Three-dimensional flow effects also cause stall delay at the root; this is typically modeled via \textit{a dynamic stall model} and associated time constant;
\item Yawed inflow is not naturally included and typically incorporated via \textit{a wake correction}.
%\item Dynamic wake - Oye time lag constant
%\item Tower shadow constants " Potential flow / Downwind Empirical model
\end{itemize}
Next to these model uncertainties we will consider geometric uncertainties: the chord and twist distribution along the blade.

@@

%\subsection{Geometric uncertainties with NURBS curves}
The chord, twist, lift coefficient, and drag coefficient are functions of the radial distance along the blade, leading to a very high-dimensional number of uncertain parameters. In order to reduce this number, we parametrize the variation along the blade by using Non-Uniform Rational Basis Splines (NURBS) \cite{rogers2000}. In this way a limited number of control points can approximate a large variety of curves, so that the resulting number of uncertain parameters is relatively small. 
%To perturb the geometrical parameters of the turbine blade, such as the reference chord or twist. This parameterization will also be used to perturb lift-and drag polars. %Further, the set of control points and knots can be directly manipulated to control the smoothness and curvature.
NURBS have already been exploited in several aerodynamics applications, for instance, design optimization of wind-turbine blades, e.g.~\cite{Bottasso2014,Ribeiro2012}, and to parametrize chord and twist curves \cite{Echeverria2017}. In the following, we briefly outline the procedure used to parametrize a curve using NURBS.

@@

In order to parameterize a given curve, e.g.\ the blade reference chord, we choose a degree $p$ and a number of knots $X =\{r_j\}_{j=1}^n$. The knots can be uniformly spaced or chosen heuristically (for advanced knot selection approaches for knot selection see \cite{Li2005}). Next we sample $S_{ref}(r)$ at $X$ and also compute $B_{i,p}(r_j), \text{ for }i, j = 1,2, ..., n$. We then formulate a linear system and solve for the set of control points $\mathbf{c}=\{c_i\}_{i=1}^n$ as:
\begin{equation}\label{nurbs_inversion}
\mathbf{B}\mathbf{c} = \mathbf{S}_{ref},
\end{equation}
where $\mathbf{S}_{ref}\in\mathbb{R}^{n}$ is a vector containing sampled values of $S_{ref}(r)$ and $\mathbf{B}\in \mathbb{R}^{n\times n}$ is a matrix with the $j$-th row consisting of values of $n$ B-splines sampled at location $r_j$. Once the control points are obtained, the approximate reference curve $S_n(r)$ follows from \eqref{NURB_curve}. Fig.~\ref{nurbs_approx} shows a parameterization of the chord curve using $n=9$ control points and second-order ($p=2$) B-splines.
\begin{figure}[h!]
\centering
\includegraphics[trim={0cm 0cm 1.5cm 0cm},clip, scale=1.25]{nurbs_approx.eps}
\caption{Parametrization of chord using 9 control points and second-order B-splines.}
\label{nurbs_approx}
\end{figure}

@@

\subsection{Sobol analysis}
We employ the variance-based Sobol decomposition to perform global sensitivity analysis. This approach allows for quantification of the relative importance of the input parameters on a scale of $[0,1]$ known as \emph{Sobol indices}. In a Sobol analysis, we express the total variance of the output in term of contributions from individual parameters and their combinations. Consider a generic BEM model $f(\mathbf{z})$, depending on uncertain parameters $\mathbf{z}$ and returning a quantity of interest $Y$ (e.g.\ power, loads, moments):
\begin{equation}\label{nonlinear_model}
Y = f({\mathbf{z}}),
\end{equation}
where ${\mathbf{z}} = [z_1, z_2, ..., z_M]\in \mathcal{D}_{\mathbf{z}}\in \mathbb{R}^M$ is a vector of random variables (parameters). For ease of notation, we assume these input parameters to be uniformly distributed i.e. $z_i \sim \mathcal{U}(0,1)$ and the support of the input set is $\mathcal{D}_{\mathbf{z}} = [0,1]^M$ where $M$ is the total number of input parameters. The Sobol decomposition is based on the following hierarchical representation \cite{RSmith}:
\begin{equation}\label{sobol_decomp}
f(z_1, z_2, ..., z_M) = f_0+\sum_{i=1}^M f_i(z_i) + \sum_{1\leq i<j\leq M} f_{ij}(z_i,z_j) + ... + f_{1, 2, ..., M}(z_1, z_2, ..., z_M).
\end{equation}
Here the zeroth-order function $f_0$ is the mean response of $f$, the first-order univariate functions $f_i(z_i)$ quantify independent contributions due to each random input, and second-order functions $f_{i,j}(z_i,z_j)$ represent the effect of interaction between $z_i$ and $z_j$ on the response $f$. Formally, the zeroth-, first- and second-order terms are defined as:
\begin{align}\label{sobol_terms}
f_0 = \int_{\mathcal{D}_{\mathbf{z}}}f(\mathbf{z})d\mathbf{z}, \qquad  f_i(z_i)= \int_{{\mathcal{D}_{\mathbf{z}}}^{M-1}}f(\mathbf{z})d\mathbf{z}_{\sim \{i\}} - f_0,\\
f_{ij}(z_i,z_j) = \int_{{\mathcal{D}_{\mathbf{z}}}^{M-2}}f(\mathbf{z})d\mathbf{z}_{\sim \{i,j\}} - f_i(z_i) - f_j(z_j) - f_0,
\end{align}
where ${\mathcal{D}_{\mathbf{z}}}^{M-1} = [0,1]^{M-1}$ and the notation $\mathbf{z}_{\sim \{i,j\}}$ denotes the vector having all the components of $\mathbf{z}$ except in the set $\{i,j\}$. Higher-order terms are interpreted in a similar manner. Note that \eqref{sobol_decomp} is only valid for independent input parameters. We consider a second-order expansion of \eqref{sobol_decomp}, which is generally sufficient for the purpose of this article.

As an example, $Y$ can represent the power output of a wind turbine, with random inputs windspeed $(z_1)$, wind standard deviation $(z_2)$ and rotational speed $(z_3)$. $f_0$ then represents the mean power output, $f_1(z_1)$ represent the independent contribution of windspeed on the power and $f_{1,2}(z_1,z_2)$ quantifies the interactions of wind speed and wind standard deviation on power.

Sobol indices are based on a variance decomposition \cite{SOBOL2001271}. The definition of the variance is:
\begin{equation}\label{tot_var}
D = \text{Var}[f(\mathbf{z})] = \int_{\mathcal{D}_{\mathbf{z}}} f^2(\mathbf{z})d\mathbf{z} - f_0^2,
\end{equation}
which is rewritten as
%As the expansion in \eqref{sobol_decomp} is not unique, some orthogonality conditions must be imposed  \cite{Rabitz1999,} which allows to express the total variance as
\begin{equation}
D = \sum_{i=1}^M D_i + \sum_{1\leq i<j\leq M} D_{ij},
\end{equation}
where the first- and second-order partial variances are defined via decomposition \eqref{sobol_decomp} as
\begin{align}
D_i &= \int_{0}^{1} f^2_i(z_i)dz_i, & D_{ij} &= \int_{0}^{1} \int_{0}^{1} f^2_{ij}(z_i,z_j)dz_idz_j \label{partial_variance2}. 
\end{align}
The first- and second-order Sobol indices are then computed as
\begin{equation}\label{sobol_ind}
S_i = \frac{D_i}{D}, \quad S_{ij} = \frac{D_{ij}}{D}, \quad i,j=1,2, ..., M.
\end{equation}
%and the sum of all indices satisfy
%\begin{equation}
%\sum_{i=1}^M S_i + \sum_{1\leq i<j\leq M} S_{ij} = 1.
%\end{equation}
The total effect of $z_i$ on the output $Y$ can be quantified using the total sensitivity index $S_{T_i} = S_i + \sum_{j=1}^M S_{ij}$. The total sensitivity indices can be interpreted as an importance measure for the parameter $z_i$, therefore a large $S_{T_i}$ implies that $z_i$ has a strong influence on $Y$. The computation of $S_{T_i}$ requires the approximation of partial variances defined in \eqref{partial_variance2}. Using the Monte Carlo method for computing these variances can be prohibitive for computationally expensive models $Y$. On the other hand, Polynomial Chaos Expansions (PCE) based approaches provide a more efficient alternative for sensitivity analysis, and will be used here. As shown in \cite{SUDRET2008964}, one can compute the Sobol indices analytically by post-processing the PCE coefficients, whose computation will be detailed next.

\subsection{PCE-based Sobol indices computation}
A PCE approximation $f^{K}(\mathbf{z})$ to the computational model $Y=f(\mathbf{z})$ can be defined as a weighted sum of multivariate polynomials in $\mathbf{z}$ \cite{RSmith}:
\begin{equation}\label{eq:PCE}
Y= f(\mathbf{z}) \approx f^{K}(\mathbf{z}) = \sum_{|\mathbf{k}| = 0}^K w_{\mathbf{k}}\Psi_{\mathbf{k}}(\mathbf{z}),
\end{equation}
where $\mathbf{k}\in \mathbb{N}_0^M$ is multi-index with magnitude $|\mathbf{k}| = k_1+k_2 + ... + k_M$ and $\Psi_{\mathbf{k}}(\mathbf{z})$ a multivariate polynomial basis, computed using a tensor product of univariate polynomials $\psi_{k_i}^{(i)}(z_i)$
\begin{equation}
\Psi_{\mathbf{k}}(\mathbf{z}) := \prod_{i=1}^M\psi_{k_i}^{(i)}(z_i),
\end{equation}
where $k_i$ is the order of the univariate polynomial. The choice of univariate orthogonal polynomial $\psi_{k_i}$ depends on the type of the random variable $z_i$. For example, for uniform random variables the Legendre family of polynomials is used. 
%The multivariate polynomials $\Psi_{\mathbf{k}}(\mathbf{z})$ is orthogonal with respect to the joint probability distribution of random vector $\mathbf{z}$. 
The total number of terms $K$ in \eqref{eq:PCE} depends on the truncation scheme; in a standard truncation scheme  all polynomials in $M$ random variable with order less than or equal to $p$ are used. In this work, a more advanced truncation scheme known as the hyperbolic truncation scheme is used \cite{BlatmanThesis}. 
%\textcolor{red}{In this work, we use the hyperbolic truncation scheme with a $qnorm = 0.75$.}

The PCE coefficients $w_{\mathbf{k}}$ corresponding to $\Psi_{\mathbf{k}}(\mathbf{z})$ can be computed using a pseudo-spectral approach by using quadrature methods. For a small number of dimensions this is an efficient approach. However, in the current work we will encounter more than 10 dimensions and quadrature methods become  too expensive as they suffer strongly from the curse of dimensionality. We therefore focus on methods in which the sampling points can be chosen more freely and adaptively: Ordinary Least Squares (OLS) and Least-Angle Regression (LARS).
%In the current work where we have a limited computational budget, methods that minimizes the approximation error for a given number of samples are more practical. Next, we briefly describe two such methods to compute PCE coefficients based on a fixed number of model evaluations.

@@

\subsubsection{Ordinary Least Squares}
%Instead of using a projection-based method one can also use least squares based approaches to compute the coefficients \cite{Berveiller2006}. 
The PCE expansion \eqref{eq:PCE} can be expressed in terms of the following vector-matrix product
\begin{equation}\label{eq:LSM}
Y = f^{K}(\mathbf{z}) +\varepsilon_K = \sum_{|\mathbf{k}| = 0}^K w_{\mathbf{k}}\Psi_{\mathbf{k}}(\mathbf{z}) + \varepsilon_K = \mathbf{w}^T\boldsymbol{\Psi}(\mathbf{z}) +\varepsilon_K,
\end{equation} 
where $\mathbf{w} = \{w_0,w_1,...,w_K\}^T$ is the coefficient vector, $\boldsymbol{\Psi}(\mathbf{z}) = \{\Psi_0(\mathbf{z}), \Psi_1(\mathbf{z}), ..., \Psi_K(\mathbf{z})\}^T$ is the matrix containing multivariate polynomials in $\mathbf{z}$, and $\varepsilon_K$ is the approximation error. The least-squares minimization problem can then be formulated as:
\begin{equation}\label{eq:PCE_OLS}
\mathbf{\overline{w}} = \argmin_{\mathbf{w}} \mathbb{E}\left[\left(\mathbf{w}^T \boldsymbol{\Psi}(\mathbf{z}) - f(\mathbf{z})\right)^2\right].
 \end{equation}
The coefficient vector $\mathbf{\overline{w}}$ can be computed using the Ordinary Least Squares (OLS) method. Given $N$ model evaluations $\mathbf{Y} = \{f(\mathbf{z}_1), f(\mathbf{z}_2), ..., f(\mathbf{z}_N)\}$, we can obtain the coefficients as:
\begin{equation}\label{eq:OLS}
\overline{\mathbf{w}} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{Y},
\end{equation}
where $\mathbf{X}\in \mathbb{R}^{N\times K}$ is the design matrix where $\mathbf{X}_{i,j} = \Psi_{j}(\mathbf{z}_i)$, $i = 1,2, ..., N$ and $j = 0,1,2,..., K$. Although the OLS method allows to compute the coefficients using a limited number of model evaluations, it may result in a PCE model that consists of high-order interaction terms that may lead to overfitting. For many engineering problems, low-order interactions between the input variables are most important. Therefore, the least squares form in \eqref{eq:PCE_OLS} can be regularized such that the minimization results in a low-order sparse PCE model, as discussed next.

\subsubsection{Least Angle Regression}
The LARS algorithm \cite{efron2004} is one of the methods that can be applied to compute sparse PCE models \cite{BlatmanThesis}. The key idea is to find the best set of polynomials from a given candidate set that minimizes the regularized form of \eqref{eq:PCE_OLS}:
\begin{equation}\label{regularized_LS}
\mathbf{\overline{w}} = \argmin_{\mathbf{w}} \mathbb{E}\left[\left(\mathbf{w}^T \boldsymbol{\Psi}(\mathbf{z}) - f(\mathbf{z})\right)^2\right] + \lambda ||\mathbf{w}||_1.
\end{equation}
The LARS algorithm in the context of PCE can be summarized as follows. We start with all PCE coefficients for the candidate polynomials set to zero, and find the polynomial, say $\Psi_1(\mathbf{w})$ most correlated with the model evaluations. This polynomial enters the active set. Next, we take the largest step in the direction of $\Psi_1(\mathbf{w})$ until some other polynomial $\Psi_2(\mathbf{w})$ from the candidate set, has as much correlation with the current residual. The second polynomial is now also added to the active set. In the next step, we move both the coefficients in the active set towards their least-square values until a third polynomial from the candidate set shows as much correlation as the current PCE model based on two polynomials shows with the model evaluations. The number of iterations is given by the minimum of $\{K,N\}$. After every iteration an a-posteriori error, for instance the Leave-One-Out (LOO) cross-validation error, is computed. The active set with the smallest LOO is then chosen as the best sparse PCE model.

@@

%\subsection{Sobol index computation}

@@

%\subsection{NM80}
\subsection{Operating conditions}
As a reference wind turbine, we consider the 2MW NM80 onshore turbine from the DANAERO project \cite{Troldborg2013} with a blade radius of 38.8 m and a hub height of 57.19 m with the mean wind speed at hub height set as 6.1m/s. Furthermore, the RPM is set at 12.3$/\text{min.}$, the blade pitch angle at 0.15 $\deg$ and the turbine yaw angle is zero.

\subsection{Choice of uncertain inputs}
The data for lift ($C_l$) and drag ($C_d$) polars are available at four locations along the blade radius: at 11.87 m, 17.82 m, 28.97 m, and 35.53 m. The lift and drag variables at these sections are numbered Cl1 - Cl4 and Cd1 - Cd4 respectively. The reference value of these polars is obtained from the wind-tunnel experiment with 3D corrections. Random samples of chord, twist, lift- and drag-polars are obtained by perturbing the control points with a uniformly distributed random variable.
